---
title: "Lab 06 - Load a Shapefile"
author: "Jesse Lecy"
date: "Sunday, February 22, 2015"
output: pdf_document
---

So far we have examined geographic point data using the `ggmap` package. We have created a base map, which is a simple image tile, in order to provide context. The point data is plotted over the top of the map on an x-y axis using latitude and longitude coordinates. The point data may have been generated by geocoding address data (translating street addresses to latitude and longitude coordinates). We might also summarize the point data by using heat maps or contour maps.

We will now start working with tangible map assets. Instead of using an image of a map as background, we will learn to manipulate elements of a map file like administrative boundaries (states, census tracts, etc.), geographic features (rivers, roads, etc.), and points (city centroids, for example). These map elements collectively are known as TIGER **shapefiles**.

According to the US Census Bureau:

"In preparation for the 1990 Census, the Census Bureau, in collaboration with the U.S. Geological Survey, developed the first nationwide digital map of the U.S., Puerto Rico and other territories called the Topologically Integrated Geographic Encoding and Referencing (TIGER) database. As a national digital map, TIGER contains all of the geographic features - such as roads, railroads, rivers, and legal and statistical geographic boundaries - that are necessary to support the Census Bureau's data collection and dissemination programs."

Shapefiles are collections of polygons, lines and points that represent elements of a map. Each element can be manipulated separately. For example, we might have a shapefile that contains the 48 continental US states, but we want to focus the analysis on the midwestern states. We can extract the elements related to the
midwest and construct a separate map of only those elements. 

Each element will have associated metadata. Each specific state has a geographic land area, a certain portion of the state covered by water, a unique geographic ID, and a centroid (the point representing the geographic center of the polygon). Shapefiles specifically are collections of polygons (or lines and points), and their associated metadata.

The most common use of shapefiles in this class will be to create choropleth maps, maps where administrative units are shaded according to values of a specific variable of interest like income, race, or employment. In this lab we will learn how to load and manipulate shapefiles. In the subsequent labs we will learn how to merge census data with map elements to create choropleth maps.

# Downloading Shapefiles from the Census

The shapefiles that we will use for this class are generated by the US Census Bureau and available for download from:

https://www.census.gov/geo/maps-data/data/tiger-line.html

When using shapefiles for map-making you will need to first select the appropriate units of anaysis. The Census provides shapefiles for states, counties, and census tracts, among others. You will download the appropriate files using the query tool available on the website.

The download query will generate a zipped folder that contains all of the necessary map files. You will need to create a folder for these shapefile files and unzip the contents into the folder.

You will notice that the download has generated up to five files, all with the same name but different file extensions .dbf, .shp, .prj, and .shx. All of these individual files collective comprise a shapefile, so make sure you keep them all together in the same directory.

# Loading Shapefiles in R

We will be using the package `maptools` and the function `readShapePoly()` for polygon shapefiles like census tracts.

Let's take a look at an example using the census tracts shapefile for Pennsylvania:

```{r}

# install.packages( "maptools" )

library( maptools )

setwd("~/Penn Census Tracts")


penn <- readShapePoly(
                        fn="tl_2010_42_tract10",
                        proj4string=CRS("+proj=longlat +datum=WGS84")
                     )


plot( penn, border="blue", axes=TRUE )

```

There are only two arguments to worry about with the `readShapePoly()` function:

Argument     | Value
-------------|-----------------------------------
fn=          | The file name without an extension
proj4string= | The map projection, use *CRS("+proj=longlat +datum=WGS84")*

Recall from above that a shapefile consists of five separate files with identical names but different file extensions. The filename argument *fn=* asks for the generic name of the shapefile, but you do not include an extension because the function will grab all of the necessary files with the same name and different extensions.

In order to produce two-dimensional maps on Cartesian x-y coordinates the spherical elements of a globe must be flattened into an x-y plane. This process is called "projecting" a map. For this class we will use the 1984 World Global Systems convention using a latitude and longitude coordinate system. You will always use the value *CRS("+proj=longlat +datum=WGS84")* for this argument.

If the shapefiles were loaded correctly you should be able to draw the map with the *plot* function.

```{r}
class( penn )

length( penn )

names( penn )
```

With shapefiles it is important to note that they contain two primary components. There is a database that contains all of the coordinates necessary to draw the polygons for each element. There is also a dataframe that contains the meta-data for each element. The `class()` function is useful to confirm that the shapefile was loaded correctly and is stored as a spatial polygon. 

The `length()` function will tell you how many elements are contained within the map. For example, a shapefile containing the states would have length 50. In this case there are 3,218 separate census tracts within the state of Pennsylvania.

The meta-data file is important for a couple of reasons. First, it contains some useful information such as the land area of a polygon (we will use this to generate population density measures in a subsequent lab). More importantly, we will add additional information to the maps by merging census data to this shapefile dataframe. It is helpful to see what variables might be included as shapefile meta-data:


Name          |  Value
--------------|-------
"STATEFP10"   | State fip
"COUNTYFP10"  | County fip
"TRACTCE10"   | Tract ID in string format
"GEOID10"     | State fip + county fip + tract ID, **used for merges**
"NAME10"      | Alternative format of tract ID for labeling maps
"NAMELSAD10"  | Alternative format of tract ID in readable form
"MTFCC10"     | Ignore this field
"FUNCSTAT10"  | Ignore this field
"ALAND10"     | Area of polygon covered by land
"AWATER10"    | Area of polygon covered by water
"INTPTLAT10"  | Centroid coordinate
"INTPTLON10"  | Centroid coordinate

# The Geographic ID Variable

The US goverment has standardized a coding system for state administrative units. These standards are released as the Federal Information Processing Standard (FIPS) codes. States, counties, and census tracts each have their own codes.

The FIPS codes have the following structure:

> GEO ID Codes:   SS-CCC-TTTT.TT

> SS = State FIPS

> CCC = County FIPS

> TTTTTT = Census Tract ID

Census tracts are designed as nested administrative units where states are comprised of counties, and counties are comprised of census tracts. Each element has a numbered ID.

In Pittsburgh, for example, one census tract might have the following codes: 42=Pennsylvania, 003=Allegheny County, 0001.02=Tract 1.b.

States have FIPS codes that range from 1-50. Each state has a list of counties that starts with the number 001 and goes through the total number of counties in the state. As a result, there are approximately 50 counties with the ID of 001. In order to identify a unique county, you need the state-county pair. The ID will look something like 42-001.

Similarly, each county in a state has a census tract with the ID 0001. To uniquely identify a census tract you need to know the state, county, and tract IDs. 

Since census tracts each have approximately 4,000 individuals, when a tract grows large enough it might be split into two separate tracts. When then happens, a sub-ID numbering system is used. If tract 0001 is split, for example, it will become 0001.01 and 0001.02.

As a result, a full census tract ID will look something like:  42-001-0001.02, or just 42001000102.

Census tracts are themselves comprised of block units, which are comprised of blocks, but the census generally does not release data that is disaggregated below the tract level for privacy reasons. As a result, we will use census tracts for most of the analysis in this class.


**Pittsburgh, PA (Metropolitan Statistical Area) (38300)**

County Name | County FIPS
------------|------------
Allegheny, PA | [42003]
Armstrong, PA | [42005]
Beaver, PA | [42007]
Butler, PA | [42019]
Fayette, PA | [42051]
Washington, PA | [42125]
Westmoreland, PA | [42129]


You can create a smaller map of a specific county with a subset operation:

```{r}

# Allegheny County FIPS = 003

allegheny <- penn[ as.character(penn$COUNTYFP10) == "003", ]

plot( allegheny, border="indianred3", col="light gray" )

```

For the demographics portion of your final project you will need to make some maps of your chosen MSA. Each MSA consists of a collection of several counties, so in order to produce a map of an MSA you first need to select the relevant elements. Again, apply the subset function but this time use a set of counties instead of a single one:


```{r}

# Use the set of County FIPS from the Pittsburgh list above.
# Note that you use ony the county codes from the SS-CCC string.

these <- c("003", "005", "007", "019", "051", "125", "129" )

pitt <- penn[ penn$COUNTYFP10 %in% these, ]

plot( pitt, border="gray", col="light blue" )

```

Note that the `%in%` operator is just a shortcut for the OR statement. It identifies census tracts that are in Allegheny County OR Armstrong County OR Beaver County etc. It will return a logical selector vector similar to the other subset operations we have done.


```{r}

head( penn$COUNTYFP10 %in% these, 500 ) 

```


# Zooming

You can zoom into specific portions of the shapefiles by defining a plotting box with the xlim and ylim arguments. 

```{r}

lim <- NULL

lim$x <- c( -80.10705, -79.82922 )

lim$y <- c( 40.31530, 40.52821 )

plot(pitt, border="gray", xlim=lim$x, ylim=lim$y )

# Alternatively try:    lim <- locator(2)

```

A clever way to select your plot box is using the `locator()` function. It allows you to click on the top-left and bottom-right corner of a zoom box, and it will automatically generate the coordinates of the box.

# Adding points

If you want to add a set of points to the plotted shapefiles, you can use the `points()' function. Similar to `geom_point()` it layers points over an existing map.

```{r}

library( ggmap )

cmu <- geocode( "carnegie mellon university" )

plot( pitt, col="gray95", border="gray", xlim=lim$x, ylim=lim$y )

points( cmu, col="red4", pch=19, cex=2 )

text( cmu, "Carnegie Mellon", col="red4", pos=3 )


```



